% lecture notes by Umut Özer
% course: st
\lhead{Lecture 15: February 21}

Today we will look at how another pair of operators transforms.

\subsection{\texorpdfstring{$T(z) \normalorder{e^{i k_{\mu} X^{\mu}(w)}}$}{An Important} OPE}%
\label{sub:an_important_ope}

Our composite field is $\normalorder{\exp(i k_{\mu} X^{\mu}(w))}$, where $k_{\mu}$ is some constant spacetime (i.e.~target space) momentum vector. This will be a fundamental ingredient to any scattering amplitude, where $k_{\mu}$ is the centre-of-mass momentum of the string.

It is useful to write this
\begin{equation}
  T(z) \normalorder{e^{i k \cdot X(w)}} = T(z) \sum_{k \geq 0} \frac{i^n}{n!} k_{\mu_1} \dots k_{\mu_n} \normalorder{X^{\mu_1}(w) \dots X^{\mu_n}(w)},
\end{equation}
where $T(z) = -\frac{1}{\alpha'} \normalorder{\partial X^{\mu} (z) \cdot \partial X_{\mu}(z)}$.
In the free theory, we can work this out exactly using Wick's theorem.
The normal ordering reminds us that there are no internal contractions, only ones between the $\partial X$'s in $T$ and the $X$'s in $e^{i k \cdot X}$.

We shall make use of the OPE \eqref{eq:14-pxxope}. There are two types of contribution to the OPE:
\begin{description}
  \item[Single Contractions:] These contribute
    \begin{equation}
      \wick{-\frac{2}{\alpha'} \normalorder{\partial X (z) \cdot \c2{\partial X(z)}} \sum_{n \geq 0} \frac{i^n}{n!} k_{\mu_1} \dots k_{\mu_n} \normalorder{\c2 X^{\mu_1} (w) \dots X^{\mu_n}(w)}}
    \end{equation}
    where the factor $2$ comes from the two ways to contract any given $X^{\mu_i}$ with any of the two $\partial X$.
    Of course, we get $n$ such terms, one for each $X^{\mu_i}$:
    \begin{align}
      \dots &= \sum_{n > 0} \frac{i^n}{n!} n (k \cdot X(w))^{n-1} k_{\nu} \left( \frac{\partial X^{\nu} (w)}{ z - w} \right) \\
	    &= \sum_{n \geq 0} \frac{i^n}{n!} (k \cdot X(w))^n i k_{\nu} \left( \frac{\partial X^{\nu}(w)}{z - w} \right) \\
	    &= \frac{1}{z - w} \partial \left( e^{i k \cdot X(w)} \right).
    \end{align}
  \item[Double Contractions:] These contribute
    \begin{align}
      \begin{split}
	&\wick{ -\frac{1}{\alpha'} \normalorder{\partial \c1 X \cdot \partial \c2 X(z)} \sum_{(i, j)} \sum_{n \geq 0} \frac{i^n}{n!} k_{\mu_1} \dots k_{\mu_i} \dots k_{\mu_j} \dots k_{\mu_n} \normalorder{X^{\mu_1} (w) \dots \c1 X^{\mu_i}(w) \dots \c2 X^{\mu_j}(w) \dots X^{\mu_n}(w)} } \\
	&= -\frac{1}{\alpha'} \sum_{n \geq 2} k_{\mu_2} \dots k_{\mu_{n-1}} \frac{i^n}{n!} n (n-1) X^{\mu_2}(w) \dots X^{\mu_n} \left( -\frac{\alpha'}{2} \right)^2 \frac{k^2}{(z - w)^2}
      \end{split} \\
      &= -\frac{\alpha'}{4} \frac{k^2}{(z - w)^2} \normalorder{\sum_{n \geq 2} (k \cdot X(w))^{n-2}} i^2 i^{n-2} \frac{n!}{n! (n-2)!} \\
      &= \frac{\alpha'}{4} \frac{k^2}{(z - w)^2} \normalorder{e^{i k \cdot X(w)}}
    \end{align}
    \begin{leftbar}
      This is the second worst OPE you will have to do in this course.
    \end{leftbar}
\end{description}

This gives:
\begin{equation}
  T(z) \normalorder{e^{i k \cdot X (w)}} = \left( \frac{\alpha' k^2 / 4}{(z - w)^2}  + \frac{\partial}{z - w}\right) e^{i k \cdot X(w)} + \dots.
\end{equation}
From this we read off that $h = \frac{\alpha' k^2}{4}$.
It is easy to see that a similar result holds for $ \overline{T}{}(\overline{z}{}) e^{i k \cdot \overline{X}{} (\overline{w}{})} $ and more generally the operator $ \normalorder{ \exp(i k_{\mu} X^{\mu}(w, \overline{w}{}))} $ has conformal weight
\begin{equation}
  \boxed{(h, \overline{h}{}) = \left( \frac{\alpha' k^2}{4}, \frac{\alpha' k^2}{4} \right)}
\end{equation}

There is a sense in which this is a purely quantum mechanically result since our string coupling $\alpha'$ appears explicitly in exactly the same way $\hbar$ would appear.
The other thing to note is that these numbers depend on $k^2$.

\subsection{The \texorpdfstring{$T(z) T(w)$}{TT} OPE and the Virasoro Algebra}%
\label{sub:the_tt_stress_ope_and_the_virasoro_algebra}

This is basically the one field we cannot do without in a conformal field theory, so if it does not transform in the correct way we have a problem.
We use
\begin{equation}
  T(z) = -\frac{1}{\alpha'} \normalorder{ \partial X \cdot \partial X(z)} \qquad \text{and} \qquad 
  \partial X^{\mu}(z) \partial X^{\nu}(w) = -\frac{\alpha'}{2} \frac{\eta^{\mu\nu}}{(z - w)^2} + \dots
\end{equation}
\begin{align}
  \begin{split}
    T(z) T(w) &= 4 \left( -\frac{1}{\alpha'} \right) \wick{\normalorder{ \partial X^{\mu}  \partial \c X_{\mu}(z)} \normalorder{\partial \c X^{\nu} \partial X_{\nu} (w)}} \\
	      &\qquad {} + 2 \left( - \frac{1}{\alpha'} \right)^2 \wick{ \normalorder{\partial \c2 X_{\mu} \partial \c1 X^{\mu}(z) } \normalorder{\partial \c2 X^{\nu} \partial \c1 X_{\nu}(w)} }
  \end{split} \\
  &= -\frac{2}{\alpha'} \frac{\eta_{\mu\nu}}{(z - w)^2} \normalorder{\partial X^{\mu}(z) \partial X^{\nu}(w)} + \frac{1}{2} \frac{\delta\indices{^{\mu}_{\nu}}}{(z - w)^2} \frac{\delta\indices{^{\nu}_{\mu}}}{(z - w)^2}.
\end{align}
For the first term we can write 
\begin{equation}
  \partial X^{\mu} (z) = \partial X^{\mu}(w) + (z - w) \partial^2 X^{\mu}(w) + \dots
\end{equation}
In the second term, we recognising the dimension of spacetime as $D = \tr(\eta_{\mu\nu})$. Then we have
\begin{equation}
  T(z) T(w) = \frac{D/2}{(z - w)^4} - \frac{2}{\alpha'} \frac{1}{(z - w)^2} \normalorder{\partial X_{\mu} \partial X^{\mu} (w)} - \frac{2}{\alpha'} \frac{1}{(z - w)} \normalorder{\partial X_{\mu} \partial^2 X^{\mu}(w)}
\end{equation}
\begin{equation}
  \label{eq:tt-ope}
  \boxed{T(z) T(w) = \frac{D / 2}{(z - w)^4} + \frac{2}{(z - w)^2} T(w) + \frac{1}{z - w} \partial T(w) + \dots}
\end{equation}
\begin{leftbar}
  This implicitly assumes radial ordering on the left-hand side.
\end{leftbar}
The second term gives us the weight $h = 2$. However, the first term should not be there; we do not expect to see a pole of order $4$.
The stress tensor only transforms as a conformal field as we expect if $D = 0$.
It will turn out that what we have been studying so far is only part of the stress tensor.
We have forgotten about the ghosts, which will give a contribution of the stress tensor as well!
Their contribution will allow for $D$ to be a finite integer, which is therefore determined by the consistency of the theory.

\section{The Virasoro Algebra}%
\label{sec:the_virasoro_algebra}

We talked about the Witt algebra and saw that it had something to do with conformal transformations.
The Witt algebra came from purely classical consideration. The TT OPE was obtained by including quantum mechanical effects, like double contractions.

The stress tensor is a field of weight $2$ so we can expand it in terms of Laurent modes $L_n$:
\begin{equation}
  T(z) = \sum_n L_n z^{-n-2}.
\end{equation}
This may be inverted to give 
\begin{equation}
  L_n = \oint_{z = 0} \frac{\dd[]{z}}{2 \pi i } z^{n+1} T(z)
\end{equation}
and similarly for the tensor $\overline{T}{}(\overline{z}{})$ and its modes $\overline{L}{}_n$.
